---
title: "DEA of the top genes with high TD score"
author: "Julia Lienard"
format: html
editor: visual
---

## Background

The list of genes with TD score (based on NIC and NNC transcripts) differences between the samples was extracted in the quarto document 04_visualization_TD_differences_FLNC.qmd and is used here to analyzed the GO term potential enrichment of the top genes.

## Loading data

```{r}
library(readr)
HighTD_genes <- read_csv("../../../../analysis/01_NIC_NNC_ratio_analysis/04_visualization_TD_differences_FLNC/TD_highTDscore_min5FLNC.csv")
```

```{r}
library(dplyr)
library(tidyr)
```

## Load list of all expressed genes

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# Load list of all expressed genes
AllExpGenes <- read_delim("../../../../analysis/01_NIC_NNC_ratio_analysis/02_Common_genes_classificationSQ_noRound_FLNC/02_output/bc01bc02common_genes_RBN_diff_FLNC.txt", delim = "\t")

AllExpGenes <- AllExpGenes %>%
  tidyr::drop_na() %>% 
  dplyr::pull(associated_gene) # n = 18179
```

## Extract 10% of top genes with different TD score

```{r}
# STRESS SAMPLE
top_Stress <- HighTD_genes %>%
  filter(Greater_in_bc01 == "yes") %>%
  tidyr::drop_na() 

TOT_Stress <- as.numeric(nrow(top_Stress))

top_Stress <- top_Stress %>%
  slice_max(order_by = RBN_Difference, n = round(TOT_Stress * 0.1)) # take 10% of top genes with TD score difference (higher in Stress

top_Stress_list <- top_Stress %>% pull(associated_gene)

# CONTROL sample
top_Control <- HighTD_genes %>%
  filter(Greater_in_bc01 == "no") %>%
  tidyr::drop_na() 

TOT_control <- as.numeric(nrow(top_Control))

top_Control <- top_Control %>% 
  slice_min(order_by = RBN_Difference, n = round(TOT_control * 0.1))

top_Control_list <- top_Control %>% pull(associated_gene)
```

```{r}
# Extraction top 10 genes with highest score for supplementary figure thesis
top_10Stress <- top_Stress %>%
  slice_max(order_by = RBN_Difference, n = 10) %>%
  dplyr::select(associated_gene,RBN_Difference)

top_10Control <- top_Control %>% 
  slice_min(order_by = RBN_Difference, n = 10) %>%
  dplyr::select(associated_gene,RBN_Difference)

# export

write.csv(top_10Control, "../../../../analysis/01_NIC_NNC_ratio_analysis/03_NIC_NNC_ratio_using_R_FLNC/top10_highTD_control.csv")

write.csv(top_10Stress, "../../../../analysis/01_NIC_NNC_ratio_analysis/03_NIC_NNC_ratio_using_R_FLNC/top10_highTD_stress.csv")
```

## Perform GO enrichment analysis

### Stress sample

```{r}
ego_top_Stress <- enrichGO(
gene = top_Stress_list,
universe = AllExpGenes,
OrgDb = org.Hs.eg.db,
keyType = "SYMBOL", # or "ENSEMBL" if using Ensembl ID
ont = "BP", # "ALL" or "BP", "MF", "CC"
pAdjustMethod = "BH",
pvalueCutoff = 0.05,
qvalueCutoff = 0.2,
readable = TRUE
)
head(ego_top_Stress)
```

```{r}
library(ggplot2)
plot_ego_top_Stress <- dotplot(ego_top_Stress) + 
  labs(title = "Stress Sample")
plot_ego_top_Stress 
```

### Control sample

```{r}
ego_top_Control <- enrichGO(
gene = top_Control_list,
universe = AllExpGenes,
OrgDb = org.Hs.eg.db,
keyType = "SYMBOL", # or "ENSEMBL" if using Ensembl ID
ont = "BP", # "ALL" or "BP", "MF", "CC"
pAdjustMethod = "BH",
pvalueCutoff = 0.05,
qvalueCutoff = 0.2,
readable = TRUE
)
head(ego_top_Control)
```

## Extract 10% of all genes with highest TD scores

```{r}
TOT_genes <- as.numeric(nrow(HighTD_genes))

all_top10genes <- HighTD_genes %>%
  slice_max(order_by = abs(RBN_Difference), n = round(TOT_genes * 0.1)) %>% pull(associated_gene)


ego_top10genes <- enrichGO(
gene = all_top10genes,
universe = AllExpGenes,
OrgDb = org.Hs.eg.db,
keyType = "SYMBOL", # or "ENSEMBL" if using Ensembl ID
ont = "BP", # "ALL" or "BP", "MF", "CC"
pAdjustMethod = "BH",
pvalueCutoff = 0.05,
qvalueCutoff = 0.2,
readable = TRUE
)
head(ego_top10genes) # we get the same enrichment as for the Stress sample alone.
```

```{r}
library(viridisLite)

plot_ego_top10genes <- dotplot(ego_top10genes) + viridis::scale_fill_viridis() + 
  labs(title = "Both samples")
plot_ego_top10genes 
```

```{r, width = 5}
library(patchwork)

plot_ego_top_Stress / plot_ego_top10genes
```
