---
title: "Kinnex data - neoantigens isoform structural characterization"
author: "Julia Lienard"
date: 2025-02-19
date-modified: last-modified
format: pdf
editor: visual
---

The data from kinnex long read sequencing (MAS-ISO seq method) of the 2 samples of human pancreatic beta cell (endoC-BetaH1) provided by Lorenzo Pasquali's lab, were processed with the isoseq pipeline until isoseq cluster2. Mapping of the reads were done against the human reference genome GRCh38.

SqantiQC was run and the classification output files (found here: /home/jlienard/BetaCells_kinnex/\\

04_sqantiQC/02_Using_GRCh38_mapping/output) generated are used here (prior merging with correct FLNC counts using 00_Merge_SqQCclass_FLcounts.R ) to focus the analysis of genes encoding known autoantigens encoding genes during type I diabetes: list of neoantigens provided based on the literature (and editable if needed) to filter the classification files.

## Import library

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

## Loading input files

```{r}
# loading output from Sqanti-QC (GRCh38 ref genome used for mapping)

# loading bc01_class_FLNC.txt (output from 00_Merge_SqQCclass_FLcounts.R)
bc01_class <- read_delim("../../../../analysis/01_NIC_NNC_ratio_analysis/00_Merge_SqQCclass_FLcounts/bc01_class_FLNC.txt" , 
                         delim = "\t", 
                         escape_double = FALSE, 
                         trim_ws = TRUE)

# loading bc02_class_FLNC.txt
bc02_class <- read_delim("../../../../analysis/01_NIC_NNC_ratio_analysis/00_Merge_SqQCclass_FLcounts/bc02_class_FLNC.txt" , 
                         delim = "\t", 
                         escape_double = FALSE, 
                         trim_ws = TRUE)

# load neoantigens_list.txt
neoantigen <- 
  read_delim(file = "../../../../00_raw_data/neoantigens_list.txt", 
                         delim = "\t",
                         col_names = FALSE)
# names of the neoantigen-encoding genes should be the one used in the human reference annotation
```

## Calculation of log2(CPM+1) as a value of isoform expression

normalizing the read counts (CPM calculation). It is calculated in SqantiQC as follow:

> log2((FL_count/bc01_mapped_reads)\*10e6) +1)

```{r}
# sum of all FL in the classification file using R:
bc01_tot_FL_count <- sum(bc01_class$FLNC, na.rm=T) 
bc02_tot_FL_count <- sum(bc02_class$FLNC, na.rm=T)

# FLNC counts for annotated and novel genes
```

## Filter classification file based on the list of neoantigens

```{r}
# imported list transformed into a list in R:
neoantigen_list <- as.list(neoantigen[[1]]) 

# filter bc01_class for keeping only the neoantigens and have only isoforms without RT switching artefacts:
bc01neoantigen_class <- bc01_class %>% 
  filter(associated_gene %in% neoantigen_list & RTS_stage == FALSE)

bc02neoantigen_class <- bc02_class %>% 
  filter(associated_gene %in% neoantigen_list & RTS_stage == FALSE)


# adding sample origine before joining dataframes
bc01neoantigen_class$sample <- "bc01"
bc02neoantigen_class$sample <- "bc02"

# joining
all_neoantigen_class <- 
  bind_rows(bc01neoantigen_class, bc02neoantigen_class)
```

### 1) Read counts

```{r}
# Step 1: Aggregate FL_count by gene and sample
all_neoantigen_class_gp_by_FL <- all_neoantigen_class %>%
  group_by(associated_gene, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), .groups = 'drop')  
# Aggregating FL counts per gene and sample

# Step 2: Calculate CPM
all_neoantigen_class_gp_by_FL <- all_neoantigen_class_gp_by_FL %>%
  mutate(CPM = (FL_sum / case_when(
    sample == "bc01" ~ bc01_tot_FL_count,
    sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(CPM + 1)
all_neoantigen_class_gp_by_FL <- 
  all_neoantigen_class_gp_by_FL %>%
  mutate(logCPM = log2(CPM + 1))
```

```{r}
head(all_neoantigen_class_gp_by_FL, 18)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot expression for isoforms no log2 transfo
Iso_expressionCPM <- 
  ggplot(all_neoantigen_class_gp_by_FL) +
  geom_col(mapping = aes(x=sample, 
                         y=CPM, 
                         fill = sample)) +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", 
                               "bc02" = "palevioletred3")) +
  facet_grid(~associated_gene) +
  labs(title = "All Isoforms expression for neoantigen-encoding genes",
       y = "CPM", x = "Pancreatic β-cell samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

Iso_expressionCPM
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot expression for for isoform
Iso_expressionlog2TPM <- ggplot(all_neoantigen_class_gp_by_FL) +
  geom_col(mapping = aes(x=sample, y=logCPM, fill = sample)) +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", 
                               "bc02" = "palevioletred3")) +
  facet_grid(~associated_gene) +
  labs(title = "All Isoforms expression for neoantigen-encoding genes",
       y = "log2(CPM+1)", x = "Pancreatic β-cell samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

Iso_expressionlog2TPM
```

#### - percentage of each transcript by gene

```{r}
# Calculate the percentage of each transcript by gene (use total FL counts per gene and FL counts for each transcript of the same gene)

all_neoantigen_class_gp_by_FL_filt <- 
  all_neoantigen_class_gp_by_FL %>% 
  select(-CPM, -logCPM)

all_neoantigen_class_totFLgene <- 
  inner_join(all_neoantigen_class_gp_by_FL_filt, all_neoantigen_class, 
             by = c("sample", "associated_gene"))

all_neoantigen_class_totFLgene <- all_neoantigen_class_totFLgene %>%
  mutate(trans_perc = FLNC * 100 / FL_sum)
```

```{r}
#| fig-width: 7
#| fig-height: 8
#| fig-align: center
#| out-width: 100%


ggplot(data = all_neoantigen_class_totFLgene, 
                     mapping = aes(x=sample, 
                                   y=log(FLNC), 
                                   color = structural_category)) +
  geom_jitter(shape = 21) +
  facet_wrap(~associated_gene, 
             scale = "free_y") + 
  labs(title = "Isoforms found for neoantigen-encoding genes",
       y = "log(FL counts)") +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("full-splice_match" = "#6BAED6", 
                                "incomplete-splice_match" = "#FC8D59", 
                                "novel_in_catalog" = "#78C679", 
                                "novel_not_in_catalog" = "#EE6A50", 
                                "genic" = "#969696"))
```

```{r}
#| fig-width: 7
#| fig-height: 8
#| fig-align: center
#| out-width: 100%


# Create a variable with for FL groups according to the counts
all_neoantigen_class_totFLgene <- all_neoantigen_class_totFLgene %>%
  mutate(FL_group =
           case_when(
    FLNC == 1 ~ "1",
    FLNC > 1 & FL < 11 ~ "2-10",
    FLNC > 10 & FL < 51 ~"11-50",
    FLNC > 50 & FL < 101 ~"51-100",
    FLNC > 100 & FL < 501 ~"101-500",
    FLNC > 500 ~">500",
    TRUE ~ "NA"
  ),
  FL_group = factor(FL_group, 
                    levels = c("1", "2-10", "11-50", "51-100", 
                               "101-500", ">500", "NA"))
  )

# plot FL groups
library(ggplot2)
library(ggpattern)  # Required for pattern aesthetics

ggplot(data = all_neoantigen_class_totFLgene) +
  geom_bar_pattern(mapping = aes(x=FL_group, 
                                 fill = structural_category, 
                                 pattern = sample),
                   stat = "count",
                   pattern_density = 0.02,
                   pattern_spacing = 0.025) +
  facet_wrap(~associated_gene, scales = "free_y") + 
  labs(title = "Isoforms found for neoantigen-encoding genes",
       y = "transcripts counts") +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59", 
                               "novel_in_catalog" = "#78C679", 
                               "novel_not_in_catalog" = "#EE6A50", 
                               "genic" = "#969696"))  +
  scale_pattern_manual(
    values = c("bc01" = "none", "bc02" = "stripe"),
    labels = c("bc01" = "Stress", "bc02" = "Control")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.box.just = "left",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pattern = "none"),
      order = 1,
      nrow = 2
    ),
    pattern = guide_legend(
      override.aes = list(fill = "white", color = "black"),
      order = 2
    )
  )
```

```{r}
#| fig-width: 7
#| fig-height: 8
#| fig-align: center
#| out-width: 100%


library(ggplot2)
library(ggpattern)
library(scales)  # for pseudo_log

ggplot(data = all_neoantigen_class_totFLgene, 
       aes(x = FL_group,
           fill = structural_category,
           pattern = sample)) +
  geom_bar_pattern(
    position = position_dodge2(preserve = "single"),
    stat = "count",
    pattern_density = 0.1,
    pattern_spacing = 0.05,
    pattern_key_scale_factor = 0.5
  ) +
  facet_wrap(~associated_gene, scales = "free_y") +
  scale_pattern_manual(
    values = c("bc01" = "none", "bc02" = "stripe"),
    labels = c("bc01" = "Stress", "bc02" = "Control")
  ) +
  scale_fill_manual(
    values = c(
      "full-splice_match" = "#6BAED6", 
      "incomplete-splice_match" = "#FC8D59", 
      "novel_in_catalog" = "#78C679", 
      "novel_not_in_catalog" = "#EE6A50", 
      "genic" = "#969696"
    ),
    labels = c(
      "full-splice_match" = "FSM", 
      "incomplete-splice_match" = "ISM", 
      "novel_in_catalog" = "NIC", 
      "novel_not_in_catalog" = "NNC", 
      "genic" = "Genic"
    )
  ) +
  scale_y_continuous(trans = "pseudo_log", 
                     breaks = c(1, 5, 10, 50, 100, 500, 1000)) +
  labs(
    title = "Isoforms Found for Neoantigen-Encoding Genes",
    y = "Transcript Counts (Pseudo-log Scale)",
    x = "Nb FL Reads per Transcript",
    fill = "Structural Category",
    pattern = "Sample"
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pattern = "none"),
      order = 1
    ),
    pattern = guide_legend(
      override.aes = list(fill = "white", color = "black"),
      order = 2
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.box.just = "left",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )

```

```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center
#| out-width: 100%


library(ggplot2)
library(ggpattern)
library(scales)  # for pseudo_log

ggplot(data = filter(all_neoantigen_class_totFLgene, 
                     sample == "bc02"), 
       aes(x = FL_group,
           fill = structural_category,
           pattern = sample)) +
  geom_bar_pattern(
    position = position_dodge2(preserve = "single"),
    stat = "count",
    pattern_density = 0.1,
    pattern_spacing = 0.05,
    pattern_key_scale_factor = 0.5
  ) +
  facet_wrap(~associated_gene, scales = "free_y") +
  scale_pattern_manual(
    values = c("bc02" = "none", "bc01" = "stripe"),
    labels = c("bc01" = "Stress", "bc02" = "Control")
  ) +
  scale_fill_manual(
    values = c(
      "full-splice_match" = "#6BAED6", 
      "incomplete-splice_match" = "#FC8D59", 
      "novel_in_catalog" = "#78C679", 
      "novel_not_in_catalog" = "#EE6A50", 
      "genic" = "#969696"
    ),
    labels = c(
      "full-splice_match" = "FSM", 
      "incomplete-splice_match" = "ISM", 
      "novel_in_catalog" = "NIC", 
      "novel_not_in_catalog" = "NNC", 
      "genic" = "Genic"
    )
  ) +
  scale_y_continuous(trans = "pseudo_log", 
                     breaks = c(1, 5, 10, 50, 100, 500, 1000)) +
  labs(
    title = "Isoforms Found for Neoantigen-Encoding Genes",
    subtitle = "Control sample",
    y = "Transcript Counts (Pseudo-log Scale)",
    x = "Nb FL Reads per Transcript",
    fill = "Structural Category",
    pattern = "Sample"
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pattern = "none"),
      order = 1,
      nrow = 5
    ),
    pattern = guide_legend(
      override.aes = list(fill = "white", color = "black"),
      order = 2
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.box = "vertical",
    legend.box.just = "left",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )

```

### 2) FL and structural category: Group by gene and sample

Creating a sub-dataframe where data are grouped by gene, `structural_category` and `sample`, before making the calculation of CPM and log2(CPM + 1).

```{r}
library(dplyr)

# Step 1: Aggregate FL_count by gene, structural_category, and sample
all_neoantigen_class_gp_by_gene <- all_neoantigen_class %>%
  group_by(associated_gene, structural_category, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), 
            .groups = 'drop')  
# Aggregating FL counts per gene, structural category, and sample

# Step 2: Calculate TPM
all_neoantigen_class_gp_by_gene <- all_neoantigen_class_gp_by_gene %>%
  mutate(CPM = (FL_sum / case_when(sample == "bc01" ~ bc01_tot_FL_count,
                                   sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(TPM + 1)
all_neoantigen_class_gp_by_gene <- all_neoantigen_class_gp_by_gene %>%
  mutate(logCPM = log2(CPM + 1))
```

```{r}
head(all_neoantigen_class_gp_by_gene, 15)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot expression for FSM
FSM_expression <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                structural_category == "full-splice_match")) +
  geom_col(mapping = aes(x=sample, y=logCPM, fill = sample)) +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", 
                               "bc02" = "palevioletred3")) +
  facet_grid(~associated_gene) +
  labs(title = "FSM Isoform FL_TPM normalized for neoantigen-encoding genes",
      y = "log2(CPM+1)", x = "Pancreatic β-celll samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

FSM_expression
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot expression for FSM no log2 transfo
FSM_expressionCPM <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                   structural_category == "full-splice_match")) +
  geom_col(mapping = aes(x=sample, 
                         y=CPM, 
                         fill = sample)) +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", 
                               "bc02" = "palevioletred3")) +
  facet_grid(~associated_gene) +
  labs(title = "FSM Isoform expression for neoantigen-encoding genes",
       y = "CPM", x = "Pancreatic β-cell samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

FSM_expressionCPM

#ggsave("neoantigen_FSM_expressionTPM.png", FSM_expressionTPM)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot expression for ISM
ISM_expression <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                structural_category == "incomplete-splice_match")) +
  geom_col(mapping = aes(x=sample, 
                         y=logCPM, 
                         fill = sample)) +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", "bc02" = "palevioletred3")) +
  facet_grid(~associated_gene) +
  labs(title = "ISM Isoform Isoform normalized read counts for neoantigen-encoding genes",
       y = "log2(CPM+1)", x = "Pancreatic β-cell samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

ISM_expression
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

FSM_ISM_expression <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                    structural_category %in% c("full-splice_match", "incomplete-splice_match"))) +
  geom_col(mapping = aes(x = sample , 
                         y = logCPM, 
                         fill = structural_category), 
           position = position_dodge(width = 0.8)) +  # Dodge bars by sample within each structural category
  scale_fill_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59")) +
  facet_grid(~associated_gene) +  # Separate plots per gene
  labs(title = "FSM and ISM Isoform normalized read counts for neoantigen-encoding genes",
       y = "log2(CPM + 1)", x = "Pancreatic β-cell samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

FSM_ISM_expression

```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# other way of plotting FSM and ISM isoform expression - no log2 transfo
FSM_ISM_expressionv2 <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                    structural_category %in% c("full-splice_match", "incomplete-splice_match"))) +
  geom_col(mapping = aes(x = sample , 
                         y = CPM, 
                         fill = structural_category), 
           position = position_dodge(width = 0.8)) +  
  # Dodge bars by sample within each structural category
  scale_fill_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59")) +
  facet_grid(~associated_gene) +  # Separate plots per gene
  labs(title = "FSM and ISM Isoform normalized read counts for neoantigen-encoding genes",
       y = "CPM", 
       x = " Pancreatic β-celll samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1))  # Rotate x-axis labels

FSM_ISM_expressionv2
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# Plotting FSM and ISM isoform FL read counts
FSM_ISM_expressionv3 <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                    structural_category %in% c("full-splice_match", "incomplete-splice_match"))) +
  geom_col(mapping = aes(x = sample , 
                         y = FL_sum, 
                         fill = structural_category), 
           position = position_dodge(width = 0.8)) +  # Dodge bars by sample within each structural category
  scale_fill_manual(values = c("full-splice_match" = "#6BAED6", "incomplete-splice_match" = "#FC8D59")) +
  facet_grid(~associated_gene) +  # Separate plots per gene
  labs(title = "FSM and ISM Isoform FL read counts for neoantigen-encoding genes",
       y = "FL read counts", x = " Pancreatic β-cell samples") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

FSM_ISM_expressionv3
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# other way of plotting FSM and ISM isoform expression
FSM_ISM_expressionv4 <- ggplot(filter(all_neoantigen_class_gp_by_gene, 
                                    structural_category %in% c("full-splice_match", "incomplete-splice_match"))) +
  geom_col(mapping = aes(x = structural_category, 
                         y = logCPM, 
                         fill = sample), 
           position = position_dodge(width = 0.8)) +  # Dodge bars by sample within each structural category
  scale_fill_manual(values = c("bc01" = "darkseagreen3", "bc02" = "palevioletred3")) +
  facet_grid(~associated_gene) +  # Separate plots per gene
  labs(title = "FSM and ISM Isoform normalized read counts for neoantigen-encoding genes",
       y = "log2(CPM + 1)", x = "Isoform structural category") +
  scale_x_discrete(labels = c("full-splice_match" = "FSM", "incomplete-splice_match" = "ISM")) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

FSM_ISM_expressionv4
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# other way of plotting FSM and ISM isoform expression - no log2 transfo
Expression_str_categ <- ggplot(all_neoantigen_class_gp_by_gene) +
  geom_col(mapping = aes(x = sample , y = logCPM, fill = structural_category), 
           position = position_dodge(width = 0.9))  +
  facet_wrap(~associated_gene, scale = "free_y") +  # Separate plots per gene
  labs(title = "Structural Isoform expression for neoantigen-encoding genes",
       y = "log2(CPM+1)", x = " Pancreatic β-cell samples") +
  scale_fill_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59", 
                               "novel_in_catalog" = "#78C679", 
                               "novel_not_in_catalog" = "#EE6A50", 
                               "genic" = "#969696")) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom")  +
guides(
    fill = guide_legend(title = "Structural category", ncol = 2))


Expression_str_categ

#ggsave("Neontigens_expression_str_categ.png", Expression_str_categ )
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting FSM and ISM isoform expression for 3 selected genes - no log2 transfo
Expression_str_categ_filt <- ggplot(filter(all_neoantigen_class_gp_by_gene,
                                        associated_gene %in% c("INS", 
                                                               "CHGA", 
                                                               "GAD2"))) +
  geom_col(mapping = aes(x = sample , 
                         y = logCPM, 
                         fill = structural_category), 
           position = position_dodge(width = 0.9))  +
  facet_wrap(~associated_gene, scale = "free_y") +  # Separate plots per gene
  labs(title = "Structural Isoform expression for neoantigen-encoding genes",
       y = "log2(CPM+1)", 
       x = " Pancreatic β-cell samples") +
  scale_fill_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59", 
                               "novel_in_catalog" = "#78C679", 
                               "novel_not_in_catalog" = "#EE6A50", 
                               "genic" = "#969696")) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom")  +
guides(
    fill = guide_legend(title = "Structural category", 
                        ncol = 2))


Expression_str_categ_filt
```

### 3) Exons : Group by gene, structural category, sample, exons

#### - only FSM

```{r}
# Step 1: Aggregate FL_count and number of exons by gene, structural_category, sample 
all_neoantigen_class_gp_by_exon <- 
  all_neoantigen_class %>%
  group_by(associated_gene, 
           structural_category, 
           sample, exons) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), 
            .groups = 'drop')  # Aggregating FL counts per gene, structural category, sample and exons

# Step 2: Calculate TPM
all_neoantigen_class_gp_by_exon <- all_neoantigen_class_gp_by_exon %>%
  mutate(CPM = (FL_sum / case_when(sample == "bc01" ~ bc01_tot_FL_count,
                                   sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(TPM + 1)
all_neoantigen_class_gp_by_exon <- 
  all_neoantigen_class_gp_by_exon %>%
  mutate(logCPM = log2(CPM + 1))
```

```{r}
head(all_neoantigen_class_gp_by_exon, 10)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting exons
FSM_exons <- ggplot(filter(all_neoantigen_class_gp_by_exon, 
                           structural_category %in% c("full-splice_match"))) +
  geom_point(mapping = aes(x = sample, 
                           y = exons, 
                           size = logCPM), 
             shape = 1) +
  facet_wrap(~associated_gene, scales = "free_y") +
  labs(title = "Number of exons for FSM isoforms among neoantigen-encoding genes",
       y = "Number of exons", x = "Pancreatic β-cell samples", 
       size = "norm. FL read counts (log2(CPM+1)" ) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom") +
  scale_y_continuous(expand = expansion(mult = 0.15))  # Adjust margin around the y-axis


FSM_exons
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting exons
FSM_exons_filt <- ggplot(filter(all_neoantigen_class_gp_by_exon, 
                           structural_category %in% c("full-splice_match") & 
                             associated_gene %in% c("INS", 
                                                    "CHGA", 
                                                    "GAD2"))) +
  geom_point(mapping = aes(x = sample, 
                           y = exons, 
                           size = logCPM), 
             shape = 1) +
  facet_wrap(~associated_gene, 
             scales = "free_y") +
  labs(title = "Number of exons for FSM isoforms among neoantigen-encoding genes",
       y = "Number of exons", x = "Pancreatic β-cell samples", 
       size = "norm. FL read counts log2(CPM+1)" ) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom") +
  scale_y_continuous(expand = expansion(mult = 0.15))  # Adjust margin around the y-axis


FSM_exons_filt
```

#### - FSM + ISM

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting exons
FSM_ISM_exons <- ggplot(filter(all_neoantigen_class_gp_by_exon, 
                           structural_category %in% c("full-splice_match", "incomplete-splice_match"))) +
  geom_point(mapping = aes(x = sample, 
                           y = exons, 
                           size = logCPM, 
                           color = structural_category), 
             shape = 1 , 
    position = position_dodge(width = 0.5)) + # Spread points)
  facet_wrap(~associated_gene, 
             scales = "free_y") +
  labs(title = "Number of exons for FSM and ISM isoforms among neoantigen-encoding genes",
       y = "Number of exons", 
       x = "Pancreatic β-cell samples", 
       size = "norm. FL read counts (log2(CPM+1)" ) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59")) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +  # Adjust margin around the y-axis
guides(
    color = guide_legend(title = "Structural category", ncol = 1))
FSM_ISM_exons
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting exons filtered by 3 selected genes
FSM_ISM_exons <- ggplot(filter(all_neoantigen_class_gp_by_exon, 
                           structural_category %in% c("full-splice_match", "incomplete-splice_match") & associated_gene %in% c("INS", "CHGA", "GAD2"))) +
  geom_point(mapping = aes(x = sample,
                           y = exons, 
                           size = logCPM, 
                           color = structural_category), 
             shape = 1 , 
    position = position_dodge(width = 0.5)) + # Spread points)
  facet_wrap(~associated_gene, scales = "free_y") +
  labs(title = "Number of exons for FSM and ISM isoforms among neoantigen-encoding genes",
       y = "Number of exons", x = "Pancreatic β-cell samples", 
       size = "norm. FL read counts (log2(CPM+1)" ) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("full-splice_match" = "#6BAED6", 
                               "incomplete-splice_match" = "#FC8D59")) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +  # Adjust margin around the y-axis
guides(
    color = guide_legend(title = "Structural category", ncol = 1))

FSM_ISM_exons
#ggsave("Select_neoantigen_FSM_ISM_exons.png", FSM_ISM_exons)
```

### 4 ) RBN: Group by gene, category, FL count

```{r}
# Aggregate FL_count by gene, structural_category, sample 
all_neoantigen_class_gp_gene_FL <- all_neoantigen_class %>%
  group_by(associated_gene, structural_category, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), .groups = 'drop')  

# pivoting wider the dataframe
library(tidyr)
all_neoantigen_class_gp_gene_FL_wide <- 
  tidyr::pivot_wider(all_neoantigen_class_gp_gene_FL, 
                     names_from = "structural_category",
                     values_from = "FL_sum", 
                     values_fill = list(FL_sum = 0))

# Calculate TD for each sample and gene
all_neoantigen_class_gp_gene_FL_wide <- 
  all_neoantigen_class_gp_gene_FL_wide %>%
  mutate(NIC_NNC = novel_in_catalog + novel_not_in_catalog) %>%
  mutate(TD = (NIC_NNC / (NIC_NNC + `full-splice_match`))) # `and `allows the sign - in the middle of the name to be considered as an operator

```

```{r}
head(all_neoantigen_class_gp_gene_FL, 10)
```

```{r}
head(all_neoantigen_class_gp_gene_FL_wide,10)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting TD
TD_neo <- ggplot(all_neoantigen_class_gp_gene_FL_wide) +
  geom_col(mapping = aes(x = sample, 
                         y = TD, 
                         fill = sample)) + 
  facet_wrap(~associated_gene, 
             scales = "free_y") +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  labs(title = "TD for neoantigen-encoding genes", 
       x = "Pancreatic β-cell samples",  
       y = "TD score")
             
TD_neo
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting RBN v2
TD_neoV2 <- ggplot(all_neoantigen_class_gp_gene_FL_wide) +
  geom_col(mapping = aes(x = sample, 
                         y = TD, 
                         fill = sample)) + 
  facet_grid(~associated_gene) +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", 
                               "bc02" = "palevioletred3")) + 
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "TD score for major neoantigen-encoding genes",
       y = "TD score", 
       x = " Pancreatic β-cell samples")
             
TD_neoV2
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plotting RBN for 3 selected genes
TD_neo_filt <- ggplot(filter(all_neoantigen_class_gp_gene_FL_wide,
                             associated_gene %in% c("INS", "CHGA", "GAD2"))) +
  geom_col(mapping = aes(x = sample, 
                         y = TD, 
                         fill = sample)) + 
  facet_wrap(~associated_gene, 
             scales = "free_y") +
  scale_fill_manual(values = c("bc01" = "darkseagreen3", 
                               "bc02" = "palevioletred3")) +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
  labs(title = "TD score for major neoantigen-encoding genes", 
       x = "Pancreatic β-cell samples", 
       y = "TD score") + 
  theme(legend.position = "none",
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title = element_text(size = 16),
    plot.title = element_text(size = 22, face = "bold"),
    strip.text = element_text(size = 14))
             
TD_neo_filt
```

### 5) FSM subcategories

```{r}
# Step 1: filter only FSM isoforms
all_neoantigen_FSM <- all_neoantigen_class %>% 
  filter(structural_category == "full-splice_match")
  
# Step 2: Aggregate FL_count by gene, structural_category, and sample
all_neoantigen_FSM <- all_neoantigen_FSM %>%
  group_by(associated_gene, subcategory, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), 
            .groups = 'drop') 
# Aggregating FL counts per gene, subcategory, and sample

# Step 3: Calculate TPM
all_neoantigen_FSM <- all_neoantigen_FSM %>%
  mutate(CPM = (FL_sum / case_when(
    sample == "bc01" ~ bc01_tot_FL_count,
    sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(TPM + 1)
all_neoantigen_FSM <- all_neoantigen_FSM %>%
  mutate(logCPM = log2(CPM + 1))

# Step 4 : reorder subcategories for better plotting
all_neoantigen_FSM$subcategory <- 
  factor(all_neoantigen_FSM$subcategory, 
         levels = c("reference_match",
                    "alternative_5end",
                    "alternative_3end",
                    "alternative_3end5end")) 
                    # Ensure "reference_match" is first
```

```{r}
head(all_neoantigen_FSM, 10)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot
FSM_subcat <- ggplot(all_neoantigen_FSM) +
  geom_col(mapping = aes(x = sample, y = CPM, fill = subcategory)) + 
  facet_wrap(~associated_gene, scales = "free_y", nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c('reference_match'='#c4e1f2',
                               "alternative_5end"='#7ccdfc',
                               "alternative_3end"='#02314d',
                               "alternative_3end5end"='#0e5a87'
                               )) +
  labs(title = " FSM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "Transcripts (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))
FSM_subcat


# try also plotting % transcripts
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot for 3 selected genes

FSM_subcat_filt <- ggplot(filter(all_neoantigen_FSM,
                             associated_gene %in% c("INS", "CHGA", "GAD2"))) +
  geom_col(mapping = aes(x = sample, 
                         y = CPM, 
                         fill = subcategory)) + 
  facet_wrap(~associated_gene, 
             scales = "free_y", 
             nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  scale_fill_manual(values = c('reference_match'='#c4e1f2',
                               "alternative_5end"='#7ccdfc',
                               "alternative_3end"='#02314d',
                               "alternative_3end5end"='#0e5a87'
                               )) +
  labs(title = " FSM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", 
       y = "Reads (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", 
                        ncol = 2)) +
  theme(legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14))

FSM_subcat_filt

# ggsave("Select_neoantigen_FSM_subcat_filt.png", FSM_subcat_filt)
```

```{r}
# Create new column with percentages
all_neoantigen_FSM <- all_neoantigen_FSM %>%
  group_by(associated_gene, sample) %>%  # Group by gene and sample
  mutate(Total_FL_sum = sum(FL_sum),  # Compute total FL_sum per gene & sample
         perc_FSM = (FL_sum / Total_FL_sum) * 100) %>%  # Calculate percentage
  ungroup()  # Remove grouping after computation
```

```{r}
head(all_neoantigen_FSM, 10)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot % of trancripts within FSM subcategories

FSM_subcat_perc <- ggplot(all_neoantigen_FSM) +
  geom_col(mapping = aes(x = sample, 
                         y = perc_FSM, 
                         fill = subcategory)) + 
  facet_grid(~associated_gene) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c('reference_match'='#c4e1f2',
                               "alternative_5end"='#7ccdfc',
                               "alternative_3end"='#02314d',
                               "alternative_3end5end"='#0e5a87'
                               )) +
  labs(title = " FSM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "FSM transcripts (%)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))

FSM_subcat_perc
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot % of trancripts within FSM subcategories for 3 selected genes

FSM_subcat_perc_filt <- ggplot(filter(all_neoantigen_FSM,
                             associated_gene %in% c("INS", "CHGA", "GAD2"))) +
  geom_col(mapping = aes(x = sample, y = perc_FSM, fill = subcategory)) + 
  facet_grid(~associated_gene) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c('reference_match'='#c4e1f2',
                               "alternative_5end"='#7ccdfc',
                               "alternative_3end"='#02314d',
                               "alternative_3end5end"='#0e5a87'
                               )) +
  labs(title = " FSM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "FSM transcripts (%)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))

FSM_subcat_perc_filt

#ggsave("Select_neo_FSM_subcat_perc_filt.png", FSM_subcat_perc_filt)
```

### 6) ISM subcategories

```{r}
# Step 1: filter only ISM isoforms
all_neoantigen_ISM <- all_neoantigen_class %>% 
  filter(structural_category == "incomplete-splice_match")
  
# Step 2: Aggregate FL_count by gene, structural_category, and sample
all_neoantigen_ISM <- all_neoantigen_ISM %>%
  group_by(associated_gene, subcategory, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), .groups = 'drop')  
# Aggregating FL counts per gene, subcategory, and sample

# Step 3: Calculate TPM
all_neoantigen_ISM <- all_neoantigen_ISM %>%
  mutate(CPM = (FL_sum / case_when(
    sample == "bc01" ~ bc01_tot_FL_count,
    sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(TPM + 1)
all_neoantigen_ISM <- all_neoantigen_ISM %>%
  mutate(logCPM = log2(CPM + 1))

# Step 4 : reorder subcategories for better plotting
all_neoantigen_ISM$subcategory <- 
  factor(all_neoantigen_ISM$subcategory, 
         levels = c("3prime_fragment",
                    "5prime_fragment",
                    "mono-exon",
                    "internal_fragment",
                    "intron_retention")) 
                      # Ensure "reference_match" is first
```

```{r}
head(all_neoantigen_ISM)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot
ISM_subcat <- ggplot(all_neoantigen_ISM) +
  geom_col(mapping = aes(x = sample, y = CPM, fill = subcategory)) + 
  facet_wrap(~associated_gene, scales = "free_y", nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("3prime_fragment" = "#c4531d",
                    "5prime_fragment" = "#e0936e",
                    "mono-exon" = "#cec2d2",
                    "internal_fragment" = "#e37744",
                    "intron_retention" = "#81eb82" )) +
  labs(title = " ISM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", 
       y = "Transcripts (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", 
                              "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))

ISM_subcat
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot for 3 selected genes
ISM_subcat_filt <- ggplot(filter(all_neoantigen_ISM,
                             associated_gene %in% c("INS", "CHGA", "GAD2"))) +
  geom_col(mapping = aes(x = sample, y = CPM, fill = subcategory)) + 
  facet_wrap(~associated_gene, scales = "free_y", nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("3prime_fragment" = "#c4531d",
                    "5prime_fragment" = "#e0936e",
                    "mono-exon" = "#cec2d2",
                    "internal_fragment" = "#e37744",
                    "intron_retention" = "#81eb82" )) +
  labs(title = " ISM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "Transcripts (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))

ISM_subcat_filt
```

```{r}
# Create new column with percentages
all_neoantigen_ISM <- all_neoantigen_ISM %>%
  group_by(associated_gene, sample) %>%  # Group by gene and sample
  mutate(Total_FL_sum = sum(FL_sum),  # Compute total FL_sum per gene & sample
         perc_ISM = (FL_sum / Total_FL_sum) * 100) %>%  # Calculate percentage
  ungroup()  # Remove grouping after computation
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot for 3 selected genes
ISM_subcat_perc_filt <- ggplot(filter(all_neoantigen_ISM,
                             associated_gene %in% c("INS", "CHGA", "GAD2"))) +
  geom_col(mapping = aes(x = sample, y = perc_ISM, fill = subcategory)) + 
  facet_wrap(~associated_gene, scales = "free_y", nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("3prime_fragment" = "#c4531d",
                    "5prime_fragment" = "#e0936e",
                    "mono-exon" = "#cec2d2",
                    "internal_fragment" = "#e37744",
                    "intron_retention" = "#81eb82" )) +
  labs(title = " ISM isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "Transcripts (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))

ISM_subcat_perc_filt

ggsave("Select_neo_ISM_subcat_perc_filt.png", ISM_subcat_perc_filt)
```

### 7) NNC subcategories

```{r}
# Step 1: filter only NNC isoforms
all_neoantigen_NNC <- all_neoantigen_class %>% filter(structural_category == "novel_not_in_catalog")
  
# Step 2: Aggregate FL_count by gene, structural_category, and sample
all_neoantigen_NNC <- all_neoantigen_NNC %>%
  group_by(associated_gene, subcategory, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), .groups = 'drop')  # Aggregating FL counts per gene, subcategory, and sample

# Step 3: Calculate TPM
all_neoantigen_NNC <- all_neoantigen_NNC %>%
  mutate(CPM = (FL_sum / case_when(sample == "bc01" ~ bc01_tot_FL_count,
                                   sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(TPM + 1)
all_neoantigen_NNC <- all_neoantigen_NNC %>%
  mutate(logCPM = log2(CPM + 1))

# Check the different existing subcategories
unique(all_neoantigen_NNC$subcategory)
```

```{r}
head(all_neoantigen_NNC)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot
NNC_subcat <- ggplot(all_neoantigen_NNC) +
  geom_col(mapping = aes(x = sample, y = CPM, fill = subcategory)) + 
  facet_wrap(~associated_gene, scales = "free_y", nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))  +
  labs(title = "Novel_not_in_catalog isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "Transcripts (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
scale_fill_manual(values=c("intron_retention"='#81eb82', 
                    "at_least_one_novel_splicesite" = "#32734d" ))

NNC_subcat
```

### 8) NIC subcategories

```{r}
# Step 1: filter only NIC isoforms
all_neoantigen_NIC <- all_neoantigen_class %>% filter(structural_category == "novel_in_catalog")
  
# Step 2: Aggregate FL_count by gene, structural_category, and sample
all_neoantigen_NIC <- all_neoantigen_NIC %>%
  group_by(associated_gene, subcategory, sample) %>%
  summarise(FL_sum = sum(FLNC, na.rm = TRUE), .groups = 'drop')  # Aggregating FL counts per gene, subcategory, and sample

# Step 3: Calculate TPM
all_neoantigen_NIC <- all_neoantigen_NIC %>%
  mutate(CPM = (FL_sum / case_when(sample == "bc01" ~ bc01_tot_FL_count,
                                   sample == "bc02" ~ bc02_tot_FL_count)) * 1e6)

# Step 3: Calculate log2(TPM + 1)
all_neoantigen_NIC <- all_neoantigen_NIC %>%
  mutate(logCPM = log2(CPM + 1))

# Check the different existing subcategories
unique(all_neoantigen_NIC$subcategory)
```

```{r}
head(all_neoantigen_NIC)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| out-width: 100%

# plot
NIC_subcat <- ggplot(all_neoantigen_NIC) +
  geom_col(mapping = aes(x = sample, y = CPM, fill = subcategory)) + 
  facet_wrap(~associated_gene, scales = "free_y", nrow = 1) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))  +
  labs(title = "Novel_in_catalog isoform subcategories among neoantigen-encoding genes",
       x = "Pancreatic β-cell samples", y = "Transcripts (counts per million)") +
  scale_x_discrete(labels = c("bc01" = "Stressed", "bc02" = "Control")) +
scale_fill_manual(values=c("combination_of_known_junctions"='#014d02',
                   "combination_of_known_splicesites"='#379637',  
                   "intron_retention"='#81eb82', 
                   "not_combination_of_annot_junctions"='#6ec091',
                   "mono-exon_by_intron_retention"="#4aaa72")) +
guides(
    fill = guide_legend(title = "Subcategory", ncol = 2))

NIC_subcat
```
